Flow-Based Cost Allocation in Linear Optimized Power Systems are non-unique but localizable

### 2020-08-14 paper/flow_based.tex
allocation fix storage shadowprice
correct shadow price correction

% general symbols
\newcommand{\state}{s_{i,t}}
\newcommand{\capacity}{S_{i}}
\newcommand{\costfactor}{\gamma_{i,t}}

@@

Now, assume a network with generators $s$, transmission lines $\ell$ and storage units $r$. Each asset $i = \{s, \ell, r\}$ adds an term for OPEX and a term for CAPEX to the total system cost $\totalcost$. The operational price for an asset $i$ is given by $o_i$, the capital price for one unit capacity expansion by $c_i$. For example the OPEX occasioned by generator $s$ is given by 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
where $\generation$ denotes its power generation at time $t$. As \cref{eq:opexgeneration} matches the form of \cref{eq:cost_decomposition} which allows us to use the above presented scheme in \cref{eq:general_scheme}. As a result we obtain 
\begin{align}
    \allocateopex[n \rightarrow s] &= 
   \operationalpricegeneration \,  \allocategeneration
\label{eq:allocate_opexGeneration_detailed}
\end{align}
which is the contribution of $\demand$ to the OPEX at generator $s$.
The quantity  
\begin{align}
 \allocategeneration = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
can be considered as the power that is produced by generator $s$ and consumed at node $n$ at time $t$. \\
In the same manner, we can follow the scheme allocate OPEX for flow $\flow$ and storage unit dispatch $\storagedispatch$. As we assume a bidirectional flow on line $\ell$, the OPEX is set proportional to the absolute value of the flow. The upper section in \cref{tab:cost_allocation_map} shows the mapping of variables to \cref{eq:general_scheme} in order to define the full OPEX allocation.

For the CAPEX allocation, it becomes crucial to look at the individual relations between operational state $\state$ and the capacity limit. All quantities for the CAPEX allocation, which we now discuss in detail in the following, are summarized in the middle section of \cref{tab:cost_allocation_map}.

\subsubsection*{Allocate CAPEX on Generators}

@@

\subsubsection*{Allocate CAPEX on Transmission Lines}

@@

\subsubsection*{Allocate CAPEX on Storages}

@@

In a simplified storage model, $\capacitystorage$ limits the storage dispatch $\storagedispatch$ and charging $\storagecharge$. Further it limits the maximal storage capacity $\storagesoc$ by a fix ratio $h_r$, denoting the maximum hours at full discharge. The storage $r$ dispatches power with efficiency $\efficiencydispatch$, charges power with efficiency $\efficiencycharge$ and preserves power from one time step $t$ to the next, $t+1$, with an efficiency of $\efficiencysoc$. In \cref{sec:zero_profit_storage_units} we formulate the mathematical details. As already shown in \cite{brown_decreasing_2020}, the total expenditures at $r$ are fully paid back by differences of the LMP at which the storage ``buys`` and ``sells`` power. Taking only the CAPEX into account the zero profit condition reduces to
\begin{align}
    \notag
    \capexstorage =& \capitalpricestorage \, \capacitystorage \\
    \notag
    =& \sum_t \left(\muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge \right) \storagedispatch \\
    &- \sum_t \lmp \incidencestorage  \storagecharge \Forall{r} 
    \label{eq:no_profit_capex_storage}
\end{align}
where $\muupperstoragedispatch$ and $\mulowerstoragedispatch$ are the shadow prices of the upper and lower capacity bound of the power dispatch of $r$ and $\mustateofcharge$ is the shadow price of the energy balance constraint.   
\Cref{eq:no_profit_capex_storage} matches the form of \cref{eq:cost_decomposition}. However when applying \cref{eq:cost_allocation} it follows naturally that to assume $\partial \storagecharge / \partial \demand = 0$, 
% that 
% \begin{align}
%     \pdv{\storagecharge}{\demand} \demand = 0 
% \end{align}
implying that the when a storage charges power, it does not supply any demand. Rather it stands with the demand on the consumer side, retrieving power from producing assets. 
This leaves us with 
\begin{align}
     \allocatecapexstorage = \left(\muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge \right) \allocatestoragedispatch
\end{align}
and the power allocation 
\begin{align}
    \allocatestoragedispatch = \pdv{\storagedispatch}{\demand} \demand
\end{align}
The latter only allocates dispatched power of storage $r$. Note that this will break \cref{eq:cost_payback} as the payments to $r$ surpass the CAPEX by an amount $\remainingcost^E_r$. 
% \begin{align}
%     \capexstorage_r + \remainingcost^E_r = \sum_n \allocatecapexstorage
% \end{align} 
However the fact that storages themselves ``buy`` power from other the assets, when charging, the cost allocation in \cref{eq:general_scheme} may be applied to the consumption of storages themselves, by replacing the demand $\demand$ with the power charge $\storagecharge$ in \cref{eq:general_scheme}. This derives payments that a storage unit $r$ has to pay to asset $i$, namely $\cost_{r \rightarrow i}$. The sum of those payments due to $r$ will the add up to $\remainingcost^E_r$.

\subsubsection*{Allocate Emission Expenditures}

@@

>>>>

% general symbols
\newcommand{\state}{s_{i,t}}
\newcommand{\capacity}{S_{i}}
\newcommand{\costfactor}{\gamma_{i,t}}
\newcommand{\capacityupper}{\bar{S}}
\newcommand{\capacitylower}{\ubar{S}}
\newcommand{\muuppernom}{\bar{\mu}^\text{nom}_{i}}
\newcommand{\mulowernom}{\ubar{\mu}^\text{nom}_{i}}

@@

Now, assume a network with generators $s$, transmission lines $\ell$ and storage units $r$. Each asset $i = \{s, \ell, r\}$ adds an term for OPEX and a term for CAPEX to the total system cost $\totalcost$.

\subsection{Allocating OPEX}

Let the operational price for an asset $i$ be given by $o_i$. Then, for example the OPEX occasioned by generator $s$ is given by 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
where $\generation$ denotes its power generation at time $t$. As \cref{eq:opexgeneration} matches the form of \cref{eq:cost_decomposition} which allows us to use the above presented scheme in \cref{eq:general_scheme}. As a result we obtain 
\begin{align}
    \allocateopex[n \rightarrow s] &= 
   \operationalpricegeneration \,  \allocategeneration
\label{eq:allocate_opexGeneration_detailed}
\end{align}
which is the contribution of $\demand$ to the OPEX at generator $s$.
The quantity  
\begin{align}
 \allocategeneration = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
can be considered as the power that is produced by generator $s$ and consumed at node $n$ at time $t$. \\
In the same manner, we can follow the scheme allocate OPEX for flow $\flow$ and storage unit dispatch $\storagedispatch$. As we assume a bidirectional flow on line $\ell$, the OPEX is set proportional to the absolute value of the flow. The upper section in \cref{tab:cost_allocation_map} shows the mapping of variables to \cref{eq:general_scheme} in order to define the full OPEX allocation.

\subsection{CAPEX Allocation}
For the CAPEX allocation, it becomes crucial to look at the individual relations between operational state $\state$ and the capacity limit. For all assets, let the capital price for one unit capacity expansion be denoted by $c_i$.  All quantities for the CAPEX allocation, which we now discuss in detail, are summarized in the middle section of \cref{tab:cost_allocation_map}.

\subsubsection{Generators}

@@

\subsubsection{Transmission Lines}

@@

\subsubsection{Storages}

@@

In a simplified storage model, $\capacitystorage$ limits the storage dispatch $\storagedispatch$ and charging $\storagecharge$. Further it limits the maximal storage capacity $\storagesoc$ by a fix ratio $h_r$, denoting the maximum hours at full discharge. The storage $r$ dispatches power with efficiency $\efficiencydispatch$, charges power with efficiency $\efficiencycharge$ and preserves power from one time step $t$ to the next, $t+1$, with an efficiency of $\efficiencysoc$. In \cref{sec:zero_profit_storage_units} we formulate the mathematical details. As already shown in \cite{brown_decreasing_2020}, the total expenditures at $r$ are fully paid back by differences of the LMP at which the storage ``buys`` and ``sells`` power. Taking only the CAPEX into account the zero profit condition reduces to
\begin{align}
    \notag
    \capexstorage =& \capitalpricestorage \, \capacitystorage \\
    \notag
    =& \sum_t \left(\muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge \right) \storagedispatch \\
    &- \sum_t \lmp \incidencestorage  \storagecharge \Forall{r} 
    \label{eq:no_profit_capex_storage}
\end{align}
where $\muupperstoragedispatch$ and $\mulowerstoragedispatch$ are the shadow prices of the upper and lower dispatch capacity bound and $\mustateofcharge$ is the shadow price of the energy balance constraint. When applying the cost allocation scheme \cref{eq:general_scheme}, it stands to reason to assume that $\partial \storagecharge / \partial \demand \cdot \demand = 0$, implying that the when a storage charges power, it does not supply any demand. Rather it stands with the demand on the consumer side, retrieving power from producing assets. 
This leaves us with 
\begin{align}
     \allocatecapexstorage = \left(\muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge \right) \allocatestoragedispatch
\end{align}
and the power allocation 
\begin{align}
    \allocatestoragedispatch = \pdv{\storagedispatch}{\demand} \demand
\end{align}
The latter only allocates dispatched power of storage $r$. Note that this will break \cref{eq:cost_payback} as the payments to $r$ surpass the CAPEX by an amount $\remainingcost^E_r$. 
% \begin{align}
%     \capexstorage_r + \remainingcost^E_r = \sum_n \allocatecapexstorage
% \end{align} 
It is crucial to note that like this, storage units perform a redistribution of money, and therefore distort the cost allocation. So, certain share of what is allocated to the CAPEX of a storage is in another time step spent by the storage in order to buy power from other assets. This effect scales with the amount of installed capacity.

It is possible to incorporate this redistribution effect into the cost allocation, by replacing the demand $\demand$ with the power charge $\storagecharge$ in \cref{eq:general_scheme}. Then, the derived payments that a storage unit $r$ has to pay to asset $i$ is given by $\cost_{r \rightarrow i}$. The sum of those payments due to $r$ will the sum up to $\remainingcost^E_r$.

% In a final step we can still allocate all cost to the demand $\demand$. However, this would require knowledge about when the power $\storagedispatch$ was initially charged: Given the amount of power $A_{r, t_{in} \rightarrow t_{out}}$ that is charged by storage $r$ at time step $t_{in}$ and discharged at $t_{out}$ ...

\subsection{Allocate Emission Expenditures}

@@

\subsection{Capacity Expansion Limit}

In more realistic setups, generators, lines or other assets can only be built up to a certain limit. This might caused by land use restrictions or social acceptance problems. %In the following give a general solution how the cost allocation may handle this.  
However, when constraining the capacity $\capacity$  for a subset $I$ of assets to an upper limit,
\begin{align}
    \capacity - \capacityupper \le 0 \resultsin{\muuppernom} \hpad \Forall{i \in I}
\label{eq:capacityexpansionmaximum}
\end{align}
the zero profit condition for asset $i$ alters and the revenues don't match the expenditures anymore. More precisely, the revenue from the shadow prices given in \cref{tab:cost_allocation_map} exceed the CAPEX $\cost_i$ by $\remainingcost_i$, thus  
\begin{align}
 \cost_i - \remainingcost_i =  \sum_t \costfactor \, \state \Forall{i}
 \label{eq:capex_generation_duality_bf2}
\end{align}
where the difference is given by 
\begin{align}
    \remainingcost_i = - \muuppernom \capacity
    \Forall{i}
\end{align}
In order to take this effect into account for the cost allocation, we update the cost factors in \cref{tab:cost_allocation_map} to

\begin{align}
    \costfactor \rightarrow \dfrac{c_i}{c_i + \muuppergenerationnom}\, \costfactor
\end{align}
This ensures that the allocated CAPEX sum up to the actual CAPEX at asset $i$ (and \cref{eq:cost_payback} holds).

\subsection{Brownfield Constraints}

In order to take already built infrastructure into account, the capacity $\capacity$ can be constrained to a minimum required capacity $\capacitylower$. Mathematically this translates to 
\begin{align}
    \capacitylower - \capacity  \le 0 \resultsin{\mulowernom} \hpad \forall{i \in I}
\label{eq:capacityexpansionminimum}
\end{align}
Again, such a setup alters the zero profit condition of asset $i$ as soon as the constraint becomes binding. 
In that case, asset $i$ does not collect enough revenue in order to match the CAPEX by a difference is given by 
\begin{align}
    \remainingcost_i = \mulowernom \capacity
    \Forall{i}
\end{align}
% In order to take this effect into account for the cost allocation, we update the cost factors in \cref{tab:cost_allocation_map} to

% \begin{align}
%     \costfactor \rightarrow \dfrac{c_i}{c_i + \muuppergenerationnom}\, \costfactor
% \end{align}
% This ensures that the allocated CAPEX sum up to the actual CAPEX at asset $i$ (and \cref{eq:cost_payback} holds).
